Optimal Fitting of Planar Curves to Prescribed Constraints

### 2013-08-27 idp/documentation/paper.tex
continued working on optimization problem section

% math commands
\DeclareMathOperator{\atan2}{atan2}
\newcommand{\apply}[2]{#1\!\left(#2\right)}
\renewcommand{\vector}[2]{\begin{pmatrix}#1\\\linebreak{}#2\end{pmatrix}}

@@

	\section{Problem Analysis}
	\label{section:problem_analysis}

@@

			Note that there may be multiple ways to describe a given curve design tool in terms of this process. Instead of modelling the selection of the most likely curve using some fairness measure, one may also propose an alternate view on the process in which the user implicitly specifies the exact curve they want in the properties they specify. For example, in a curve design tool which takes two points from the user and connects them using a straight line, one may say that the user provides the start and the end point of the curve, and the fairness measure is designed to always select the straight line as the fairest curve. Another way of looking at this process is to say that the user, by supplying \(p_0\) and \(p_1\), specified the coefficients of the linear term \(p_0 + \left(p_1 - p_0\right) \cdot t\), thus uniquely identifying the desired curve. Modelling the process in such a way that the user supplies only the minimum amount of implicit information, relying on the fairness measure for everything else, is usually closer to reality when analyzing usability of some curve design tool. In a similar fashion, one may consider shortcomings (for instance insufficient continuity) as either shortcomings of the specification language (the language does not allow the user to request \(G^2\) continuity) or as shortcomings of the fairness measure (the fairness measure does not select sufficiently smooth curves).

@@

				In existing vector graphics software, Bézier splines are edited by more or less directly modifying the coefficient points of the underlying Bézier curves. In the case of cubic Bézier curves, the first and the last coefficient points represent the curve's start and end points, while the differences between the second and the first as well as the fourth and the third coefficient points are proportional to the velocity of the curve at the start and end points, respectively. In that sense, the user can specify start end end points of the curve, as well as start and end velocities of the curve using the coefficient points. Most vector graphics tools have functions that allow preserving the \(G^1\) continuity of the spline by automatically modifying adjacent coefficient points accordingly. They usually do not ensure \(G^2\) continuity though. Some tools can also modify existing coefficient points in such a way that the curve passes through a point that is chosen by the user without adding more Bézier curves to the spline.

@@

				The issue with the fairness measure is that it always selects the uniquely determined cubic polynomial, thus not guaranteeing \(G^2\) continuity. It is difficult for humans to choose the coefficient points in such a way that \(G^2\) continuity is established, making it hard to create smooth curves if the design software does not support the user in this regard. The problem is aggravated by the fact that curvature discontinuities can be hard to spot and may thus only be discovered much later in the design process, increasing the cost of these mistakes. In case the software allows the user to request \(G^2\) continuous Bézier splines, this sacrifices some degree of freedom for each node, thus restricting the way nodes can be manipulated and in turn possibly forcing the user to place more nodes in order to get the desired result. Placing more nodes leads to some disadvantages that will be discussed further in the paragraphs dealing with specification language expressiveness.

				The specification language used to specify Bézier splines in common design tools is lacking in expressiveness. Specifically, describing curves in terms of their curvature is difficult since curvature may only be specified indirectly through the magnitude of the velocity values (higher velocity resulting in lower curvature). Unfortunately, this is also very dependent on the point and velocity coefficients at the other end of the spline segment. There is also no way to directly specify linearly changing curvature. This makes specifying curvature very tedious, and in case the winding angle of the curve segment is large enough, actually impossible, forcing the designer to use additional nodes in order to get the desired result. Using an excessive number of nodes makes it more tedious and time-consuming for the designer to create and/or modify such curves, since many nodes and handles have to be created or moved in order to change the overall curve. It also sacrifices smoothness, since it's easy to accidently introduce subtle bumps in the curve when using a large number of nodes. Another issue is that without software support for \(G^2\) continuity, using more nodes usually also introduces more points where curvature is not continuous. In cases where the curvature is an important property of the curve (for example the constant curvature of circular arcs, or the linearly changing curvature of spirals), it may be very difficult to specify these properties indirectly using the coefficient points.

@@

				Spiro splines consist of parts of the Euler spiral, which is a curve whose curvature changes linearly with the curve's arc length. The user can specify control points through which the software puts an interpolating spline. There is no way to directly specify tangent angle and/or curvature. The splines are guaranteed to have \(G^2\) continuity.

				Again, the model from section \ref{section:curve_design_process} is applied. With Spiro splines, the modelling is fairly straightforward. The user supplies a set of points for which the software computes an interpolating spline. The fairness measure chooses the segments of this spline to be parts of the Euler spiral and makes sure that there is \(G^2\) continuity and no excessive winding.

@@

				The only real issue lies in the insufficient expressiveness of the specification language. Section \ref{section:bézier_splines} discussed how the lack of ability to specify curvature using common Bézier spline design tools adversively affects usability. Many of the same points apply to Spiro spline design tools, since they allow the user to neither specify tangent angle nor curvature. These properties can only be expressed indirectly, by placing the control points in such a way that the algorithm will create a curve that has the desired properties, resulting in the usability issues described previously, such as requiring many control points to be specified even though the curve could be described much more concisely in terms of tangent angle and/or curvature. This seems especially unfortunate when considering that the spline primitive of Spiro splines, the Euler spiral, would lend itsself excellently to describing circular arcs, spirals and other shapes exhibiting constant or linearly changing curvature. Sometimes, the algorithm can be coaxed into producing the desired curve by placing just a few control points very carefully, resulting in a series of Euler spiral segments that exhibit the wanted properties. Unfortunately, this is rarely practical since it both requires a lot of time to set up and doesn't behave very well when modified. These issues are somewhat alleviated by the guaranteed \(G^2\) continuity together with the general tendency of the algorithm to produce extremely smooth curves even when many control points are involved. That way, when indirectly specifying tangent angle and/or curvature by placing many control points, it's not as easy to accidently create small bumps as with Bézier splines.

@@

			Definitions about vectors in the euclidean plane:
			\begin{align*}
				\overline{\cdot} & : \mathbb{R}^2 \rightarrow \mathbb{R}^2     & \overline{v}      & = \vector{-v_2}{v_1}                      && \text{orthogonal vector}\\
				\alpha           & : \mathbb{R}^2 \rightarrow \mathbb{R}       & \apply{\alpha}{v} & = \apply{\atan2}{v_2, v_1}                && \text{polar coordinate angle}
			\end{align*}

			Definitions about parametric curves:
			\begin{align*}
				p                & : \left[0,1\right] \rightarrow \mathbb{R}^2 &                   &                                           && \text{point}\\
				s                & : \left[0,1\right] \rightarrow \mathbb{R}   & \apply{s}{t}      & = \left\|\apply{p'}{t}\right\|            && \text{speed}\\
				l                & : \left[0,1\right] \rightarrow \mathbb{R}   & \apply{l}{t}      & = \int_{0}^{t}\apply{s}{u}\;\mathrm{d}u   && \text{covered arc length}\\
				\theta           & : \left[0,1\right] \rightarrow \mathbb{R}   & \apply{\theta}{t} & = \apply{\alpha}{\apply{p'}{t}}           && \text{direction}\\
				\kappa           & : \left[0,1\right] \rightarrow \mathbb{R}   & \apply{\kappa}{t} & = \frac{\apply{\theta'}{t}}{\apply{s}{t}} = \frac{\apply{p''}{t}^\mathrm{T}\overline{\apply{p'}{t}}}{\apply{s}{t}^3} && \text{curvature}
			\end{align*}

@@

@@

			The goal of this project is to figure out which specification language and fairness measure yield the best usability in the resulting curve design tool. In order to achieve this goal, we decided to use nonlinear optimization on polynomial splines, that is, the curves are splines built from parametric polynomial curves which are connected with \(G^2\) continuity and whose parameters are determined using nonlinear optimization on some objective function determined by the specification and the fairness measure. This approach allows us to conveniently experiment with different specification languages and fairness measures in order to assess their usability.

@@

				The specification language we propose uses differential geometric properties of curves and is thus decoupled from specifics of the underlying curve model, hopefully leading to a specification language that is easier to learn and use. The human visual system seems to be fairly insensitive to curve properties beyond the second derivative, so the specification language will focus on the properties point, direction and curvature as defined in \ref{section:parametric_plane_curves}. In order to provide the maximum flexibility, the specification language should allow to specify any combination of these three properties.

@@

				Having chosen a specification language and a fairness measure, it is time to formulate the optimization problem. Since polynomials of very high degree usually don't behave very well, an approach where multiple polynomial parametric curves are connected to a spline seems advisable. This way, the optimization domain is a vector consisting of all the coefficients of all the polynomials that make up the spline. In the following, we will denote the \(i\)th coefficient of the \(k\)th polynomial by \(c_{k,i}\), and the vector of all coefficients by \(c\).

				Since we chose to give the position of a specification item as a fraction of the curve's total arc length, we need a function mapping covered arc length to parameter values (the inverse to the function \(l\) defined in section \ref{section:parametric_plane_curves}). Unfortunately, this function isn't very well suited to be used as a term in a smooth opimization problem even for simple curves like polynomials. Thus, an additional optimization goal is to find parametric curves where this function is approximately linear. This property is equivaent to the speed of the curve being approximately constant and has additional useful consequences, such as the second derivative of the curve being orthogonal to the first derivative. Since the chosen specification language already specifies the total arc length of the curve, and the parameter runs through the unit range, the target value for the speed is simply the total arc length of the curve (\(l^*\)), leading to the following error term:
				\begin{displaymath}
					\int_{0}^{1}\left(\apply{s}{t} - l^*\right)^2\;\mathrm{d}t
				\end{displaymath}

@@

				% segmentation, separation of segmentation and specification, global positions on curve
				% give objective and constraints function

@@

		% choice of platform
		%   C#: platform independent, high-level, possibility to interface with native code, familarity, good UI frameworks (GTK#)
		%   ipopt
		%   casadi
		%   link to googlecode page
		% numeric optimization
		%   formalization of model using symbolic functions in casadi
		%   optimization using ipopt
		%   caching
		%   keeping track of disambiguation information
		%   input is a set of specifications, disambiguation, etc.
		%   result is a curve whose segmentation is no longer visible
		% user interface (include screenshots)
		%   goal: proof of concept
		%   interacts with numeric optimizer
		%   specify any combination of point, direction and curvature at positions on the curve
		%   change position of a specification
		%     in case of pointspecifications optionally moving the point along the curve accordingly
		%   insert/remove curve length at or between specifications
		%   displays deviation from set velocity, such that the user knows when a segment is over-stretched, or where the curve degree/segment count is too low
		%   loading/saving of documents
		%     specification can be serialized from/to XML

>>>>

% math commands
\DeclareMathOperator{\atan2}{atan2}
\newcommand{\apply}[2]{#1\!\left(#2\right)}
\newcommand{\vectorA}[1]{\begin{pmatrix}#1\end{pmatrix}}
\newcommand{\vectorB}[2]{\begin{pmatrix}#1\\\linebreak{}#2\end{pmatrix}}
\newcommand{\vectorC}[3]{\begin{pmatrix}#1\\\linebreak{}#2\\\linebreak{}#3\end{pmatrix}}

@@

	% TODO: maybe we don't want to mention G^n and C^n until formally introduced later, doesn't add much value to the usability discussion anyways
	\section{Problem Analysis}
	\label{section:problem_analysis}

@@

			Note that there may be multiple ways to describe a given curve design tool in terms of this process. Instead of modelling the selection of the most likely curve using some fairness measure, one may also propose an alternate view on the process in which the user implicitly specifies the exact curve they want in the properties they specify. For example, in a curve design tool which takes two points from the user and connects them using a straight line, one may say that the user provides the start and the end point of the curve, and the fairness measure is designed to always select the straight line as the fairest curve. Another way of looking at this process is to say that the user, by supplying \(p_0\) and \(p_1\), specified the coefficients of the linear term \(p_0 + \left(p_1 - p_0\right) \cdot t\), thus uniquely identifying the desired curve. Modelling the process in such a way that the user supplies only the minimum amount of implicit information, relying on the fairness measure for everything else, is usually closer to reality when analyzing usability of some curve design tool. In a similar fashion, one may consider shortcomings (for instance insufficient continuity) as either shortcomings of the specification language (the language does not allow the user to request \(\mathrm{G}^2\) continuity) or as shortcomings of the fairness measure (the fairness measure does not select sufficiently smooth curves).

@@

				In existing vector graphics software, Bézier splines are edited by more or less directly modifying the coefficient points of the underlying Bézier curves. In the case of cubic Bézier curves, the first and the last coefficient points represent the curve's start and end points, while the differences between the second and the first as well as the fourth and the third coefficient points are proportional to the velocity of the curve at the start and end points, respectively. In that sense, the user can specify start end end points of the curve, as well as start and end velocities of the curve using the coefficient points. Most vector graphics tools have functions that allow preserving the \(G^1\) continuity of the spline by automatically modifying adjacent coefficient points accordingly. They usually do not ensure \(\mathrm{G}^2\) continuity though. Some tools can also modify existing coefficient points in such a way that the curve passes through a point that is chosen by the user without adding more Bézier curves to the spline.

@@

				The issue with the fairness measure is that it always selects the uniquely determined cubic polynomial, thus not guaranteeing \(\mathrm{G}^2\) continuity. It is difficult for humans to choose the coefficient points in such a way that \(\mathrm{G}^2\) continuity is established, making it hard to create smooth curves if the design software does not support the user in this regard. The problem is aggravated by the fact that curvature discontinuities can be hard to spot and may thus only be discovered much later in the design process, increasing the cost of these mistakes. In case the software allows the user to request \(\mathrm{G}^2\) continuous Bézier splines, this sacrifices some degree of freedom for each node, thus restricting the way nodes can be manipulated and in turn possibly forcing the user to place more nodes in order to get the desired result. Placing more nodes leads to some disadvantages that will be discussed further in the paragraphs dealing with specification language expressiveness.

				The specification language used to specify Bézier splines in common design tools is lacking in expressiveness. Specifically, describing curves in terms of their curvature is difficult since curvature may only be specified indirectly through the magnitude of the velocity values (higher velocity resulting in lower curvature). Unfortunately, this is also very dependent on the point and velocity coefficients at the other end of the spline segment. There is also no way to directly specify linearly changing curvature. This makes specifying curvature very tedious, and in case the winding angle of the curve segment is large enough, actually impossible, forcing the designer to use additional nodes in order to get the desired result. Using an excessive number of nodes makes it more tedious and time-consuming for the designer to create and/or modify such curves, since many nodes and handles have to be created or moved in order to change the overall curve. It also sacrifices smoothness, since it's easy to accidently introduce subtle bumps in the curve when using a large number of nodes. Another issue is that without software support for \(\mathrm{G}^2\) continuity, using more nodes usually also introduces more points where curvature is not continuous. In cases where the curvature is an important property of the curve (for example the constant curvature of circular arcs, or the linearly changing curvature of spirals), it may be very difficult to specify these properties indirectly using the coefficient points.

@@

				Spiro splines consist of parts of the Euler spiral, which is a curve whose curvature changes linearly with the curve's arc length. The user can specify control points through which the software puts an interpolating spline. There is no way to directly specify tangent angle and/or curvature. The splines are guaranteed to have \(\mathrm{G}^2\) continuity.

				Again, the model from section \ref{section:curve_design_process} is applied. With Spiro splines, the modelling is fairly straightforward. The user supplies a set of points for which the software computes an interpolating spline. The fairness measure chooses the segments of this spline to be parts of the Euler spiral and makes sure that there is \(\mathrm{G}^2\) continuity and no excessive winding.

@@

				The only real issue lies in the insufficient expressiveness of the specification language. Section \ref{section:bézier_splines} discussed how the lack of ability to specify curvature using common Bézier spline design tools adversively affects usability. Many of the same points apply to Spiro spline design tools, since they allow the user to neither specify tangent angle nor curvature. These properties can only be expressed indirectly, by placing the control points in such a way that the algorithm will create a curve that has the desired properties, resulting in the usability issues described previously, such as requiring many control points to be specified even though the curve could be described much more concisely in terms of tangent angle and/or curvature. This seems especially unfortunate when considering that the spline primitive of Spiro splines, the Euler spiral, would lend itsself excellently to describing circular arcs, spirals and other shapes exhibiting constant or linearly changing curvature. Sometimes, the algorithm can be coaxed into producing the desired curve by placing just a few control points very carefully, resulting in a series of Euler spiral segments that exhibit the wanted properties. Unfortunately, this is rarely practical since it both requires a lot of time to set up and doesn't behave very well when modified. These issues are somewhat alleviated by the guaranteed \(\mathrm{G}^2\) continuity together with the general tendency of the algorithm to produce extremely smooth curves even when many control points are involved. That way, when indirectly specifying tangent angle and/or curvature by placing many control points, it's not as easy to accidently create small bumps as with Bézier splines.

@@

			Definitions about vectors in the euclidean plane:
			\begin{align*}
				\overline{\cdot} & : \mathbb{R}^2 \rightarrow \mathbb{R}^2     & \overline{v}      & = \vectorB{-v_2}{v_1}                      && \text{orthogonal vector}\\
				\alpha           & : \mathbb{R}^2 \rightarrow \mathbb{R}       & \apply{\alpha}{v} & = \apply{\atan2}{v_2, v_1}                && \text{polar coordinate angle}
			\end{align*}

			Definitions about parametric curves:
			\begin{align*}
				p                & : \left[0,1\right] \rightarrow \mathbb{R}^2 &                   &                                           && \text{point}\\
				s                & : \left[0,1\right] \rightarrow \mathbb{R}   & \apply{s}{t}      & = \left\|\apply{p'}{t}\right\|            && \text{speed}\\
				l                & : \left[0,1\right] \rightarrow \mathbb{R}   & \apply{l}{t}      & = \int_{0}^{t}\apply{s}{u}\;\mathrm{d}u   && \text{covered arc length}\\
				\theta           & : \left[0,1\right] \rightarrow \mathbb{R}   & \apply{\theta}{t} & = \apply{\alpha}{\apply{p'}{t}}           && \text{direction}\\
				\kappa           & : \left[0,1\right] \rightarrow \mathbb{R}   & \apply{\kappa}{t} & = \frac{\apply{\theta'}{t}}{\apply{s}{t}} && \text{curvature}
			\end{align*}

@@

		\subsection{Continuity}
		\label{section:continuity}

			% TODO: introduce parametric and geometric continuity

@@

			The goal of this project is to figure out which specification language and fairness measure yield the best usability in the resulting curve design tool. In order to achieve this goal, we decided to use nonlinear optimization on polynomial splines, that is, the curves are splines built from parametric polynomial curves whose parameters are determined using nonlinear optimization on some objective function determined by the specification and the fairness measure. This approach allows us to conveniently experiment with specification languages and fairness measures in order to assess their usability.

@@

				The specification language we propose uses differential geometric properties of curves and is thus decoupled from specifics of the underlying curve model, hopefully leading to a specification language that is easier to learn and use. The human visual system seems to be fairly insensitive to curve properties beyond the second derivative, so the specification language will focus on the properties point, direction and curvature as defined in \ref{section:parametric_plane_curves}. In order to provide the maximum flexibility, the specification language should allow to specify any combination of these three properties at any number of positions on the curve.

@@

				Having chosen a specification language and a fairness measure, it is time to state the optimization problem. Since polynomials of very high degree usually don't behave very well, an approach where multiple polynomial parametric curves are connected to a spline seems advisable. This way, the optimization domain is a vector consisting of all the coefficients of all the polynomials that make up the spline. In the following, we will denote the \(i\)th coefficient of the \(k\)th polynomial by \(c_{k,i}\), and the vector of all coefficients by \(c\).

				% TODO: introduce constraints and objective

				Using definitions from section \ref{section:parametric_plane_curves}...

				Given specification items ..., total arc length \(l^*\)...

				Since we chose to give the position of a specification item as a fraction of the curve's total arc length, we need the function mapping the covered arc length to the corresponding parameter values in order to state the error terms for the specification items. Unfortunately, this function isn't very well suited to be used as a term in a smooth opimization problem even for simple curves like polynomials. Additionally, we need to incorporate the specification of the total arc length of the curve into the optimization goals. In order to address these problems, the following error term is used as part of the objective function:
				\begin{displaymath}
					\apply{se}{c} = \int_{0}^{1}\left(\apply{s_c}{t} - l^*\right)^2\;\mathrm{d}t
				\end{displaymath}
				For polynomial splines with sufficient degree and segment count, this error term will be brought close to zero, making the speed approximately constant with value \(l^*\). This, in turn, makes the covered arc length function approximately linear:
				\begin{displaymath}
					\apply{l_c}{t} = \int_{0}^{t}\apply{s_c}{u}\;\mathrm{d}u \approx \int_{0}^{t}l^*\;\mathrm{d}u = l^*t
				\end{displaymath}
				This causes the inverse of the covered arc length function \(l_c^{-1}\) to be approximately linear as well, facilitating the simple statement of the error terms for the specification items. It also causes the total arc length of the curve to be equal to \(l^*\).

				As mentioned in section \ref{section:specification_language}, we are not concerned with properties beyond the second derivative of the curve. Thus, we will only require \(\mathrm{G}^2\) continuity for the whole spline. While the polynomial curve segments are smooth (\(\mathrm{C}^\infty\)), the resulting spline in general is not. The segments have to be pieced together in such a way as to ensure a certain degree of smoothness. Thus, to ensure \(\mathrm{G}^2\) continuity for the whole curve, it is sufficient to ensure \(\mathrm{G}^2\) continuity at the segment connection points. Since the constant speed requirement introduced in the previous paragraph collapses \(\mathrm{C}^2\) continuity and \(\mathrm{G}^2\) continuity, ensuring \(\mathrm{C}^2\) continuity is equivalent, yet better suited for nonlinear optimization. Thus, for each connection point (\(0 \leq k \leq n - 1\)), three constraints are added:
				\begin{align*}
					\apply{p_{c_k}}{1}   & = \apply{p_{c_{k+1}}}{0}\\
					\apply{p'_{c_k}}{1}  & = \apply{p'_{c_{k+1}}}{0}\\
					\apply{p''_{c_k}}{1} & = \apply{p''_{c_{k+1}}}{0}
				\end{align*}

@@

				% segmentation, separation of segmentation and specification, global positions on curve
				% give objective and constraints function
				% check code to see if anything was left out

@@

		% choice of platform
		%   C#: platform independent, high-level, possibility to interface with native code, familarity, good UI frameworks (GTK#)
		%   ipopt
		%   casadi
		%   link to googlecode page
		% numeric optimization
		%   trapezoid sum to approximate integrals
		%   formalization of model using symbolic functions in casadi
		%   optimization using ipopt
		%   caching
		%   keeping track of disambiguation information
		%   input is a set of specifications, disambiguation, etc.
		%   result is a curve whose segmentation is no longer visible
		% user interface (include screenshots)
		%   goal: proof of concept
		%   interacts with numeric optimizer
		%   specify any combination of point, direction and curvature at positions on the curve
		%   change position of a specification
		%     in case of pointspecifications optionally moving the point along the curve accordingly
		%   insert/remove curve length at or between specifications
		%   displays deviation from set velocity, such that the user knows when a segment is over-stretched, or where the curve degree/segment count is too low
		%   loading/saving of documents
		%     specification can be serialized from/to XML
